Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-12 paper/preprint/paper.tex
only minor changes, happy to merge

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning the
representations directly from data, guided by narrow objectives such as
predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
Bengio, and Courville 2016}). Modern advances in computing have made it
possible to provide such models with ever growing degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so, the models learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so called ``lottery tickets'' suggests that modern neural
networks can be pruned by up to 90\% while preserving their predictive
performance (\citeproc{ref-frankle2018lottery}{Frankle and Carbin 2019})
and generalizability (\citeproc{ref-morcos2019success}{Morcos et al.
2019}). Similarly, Zhang et al.
(\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are so expressive that they can fit
randomly labeled data. Thus, looking at the predictive performance, the
solutions may seem to provide compelling explanations for the data, when
in fact they are based on purely associative, semantically meaningless
patterns. This poses two related challenges. Firstly, there is no
dependable way to verify if such complex representations correspond to
meaningful and plausible explanations. Secondly, even if we could
resolve the first challenge, it remains undecided how to ensure that
models can \emph{only} learn valuable explanations.

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning
representations directly from data, guided by narrow objectives such as
predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
Bengio, and Courville 2016}). Modern advances in computing have made it
possible to provide such models with ever-growing degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so, models learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so called ``lottery ticket hypothesis'' suggests that modern
neural networks can be pruned by up to 90\% while preserving their
predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
Carbin 2019}) and generalizability
(\citeproc{ref-morcos2019success}{Morcos et al. 2019}). Similarly, Zhang
et al. (\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are so expressive that they can fit
randomly labeled data. Thus, looking at the predictive performance, the
solutions may seem to provide compelling explanations for the data, when
in fact they are based on purely associative, semantically meaningless
patterns. This poses two related challenges. Firstly, there is no
dependable way to verify if such complex representations correspond to
meaningful and plausible explanations. Secondly, even if we could
resolve the first challenge, it remains undecided how to ensure that
models can \emph{only} learn valuable explanations.
